---
title: Inspect data and ask questions
author:
- Freie Universität Berlin -- M.Sc. Landschaftsarchäologie
- Modellieren in der Landschaftsarchäologie
- Dr. Daniel Knitter
date: WS 2015/106
email: daniel.knitter@fu-berlin.de
bibliography: VS_Modellierung_Literature.bib
csl: harvard1.csl
output:
  html_document:
    toc: true
    number_sections: true
    theme: default
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
highlight: pygments
---

# Prerequisites #

**Set working directory**

```{r}
setwd("/media/daniel/homebay/Teachings/WS_2015-2016/VS_Modellieren_Landschaftsarchaeologie/")
```	

**Load data**

Get the data from xls(x) files -- load required package

Steps:

- get help
- load required package

```{r message=FALSE}
##??xls
library(gdata)
raw.data <- read.xls(xls = "./data/Fundstellen_UTM.xlsx", sheet = 1)
```

# Play with the data #

Check structure and content

```{r}
str(raw.data)
```

## Ask questions ##

### How many archaeological sites are in the data set? ###
   
```{r}
length(raw.data$Nr.)
```

**Exercise:** Show another way how to answer the question?

### How many settlements? ###

```{r}
sum(raw.data$settlement)
```

**Question:**

1. What other possibilities are there to query the amount of settlements?
2. What other question could be asked?


# Create a dataset for further analyses #

## Clean data ##

**Some mandatory things: **

- Are there missing values? 
- Do you understand every column/variable? 
- Where are problems with the data? 
- Typos? 
- What data are necessary/unnecessary?

The easiest but sometimes quite verbose way to check:

```{r}
str(raw.data)
```

Examples of more controlled queries:

```{r}
head(names(raw.data))
head(sapply(raw.data, class))
head(levels(raw.data$Epoch..site.))
tail(levels(raw.data$Site.type.), 10)
```

The "Epoch..site." variable as well as the "Site.type." variable are a mess. Fortunately, we do not need to use them, since there are already separated and precoded into binary variables.

**Questions: **

1. What does "binary variable" mean? 
2. Why is it useful? 
3. Are there other ways to code a variable?

**Exercise [for the rest of the semester...] **

Clean the messed variables yourself!

- Hint 1: try `?gsub`
- Hint 2:

```{r}
grep(pattern = "Bronze Age", x = raw.data$Epoch..site.)
```

**Question:** What do the warnings mean?

***

## What is the proportion of the different settlement types? ##

One ordinary way

```{r}
a <- raw.data[,4:7]
b <- cbind(sum(a[,1]),sum(a[,2]),sum(a[,3]),sum(a[,4]))
colnames(b) <- c("settlement","open.settlement","cave.settlement","tell.settlement")
b
c <- b/b[1]*100
c
```

One of the *apply ways

```{r}
a <- data.frame(settlement = raw.data$settlement[raw.data$settlement==1],
                open.settlement = raw.data$open.settlement[raw.data$settlement==1],
                cave.settlement = raw.data$cave.settlement[raw.data$settlement==1],
                tell.settlement = raw.data$tell.settlement[raw.data$settlement==1]
                )
sapply(a, sum)
sapply(a, function(x){sum(x)/length(x)*100})
```

The [magrittr](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) and [dplyr](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) way:

```{r message=FALSE}
library(magrittr)
library(dplyr)

## Version 1
raw.data %>%
    filter(settlement==1) %>%
    select(settlement, open.settlement, cave.settlement, tell.settlement) %>%
    summarise(settlement=sum(.[,1])/n()*100,
              open.settlement=sum(.[,2])/n()*100,
              cave.settlement=sum(.[,3])/n()*100,
              tell.settlement=sum(.[,4])/n()*100
              )

## Version 2a
raw.data %>%
    filter(settlement==1) %>%
    select(settlement, open.settlement, cave.settlement, tell.settlement) %>%
    summarise(settlements = sum(settlement),
              open.settlement = sum(open.settlement),
              cave.settlement = sum(cave.settlement),
              tell.settlement = sum(tell.settlement)
              )

## Version 2b
raw.data %>%
    filter(settlement==1) %>%
    select(settlement, open.settlement, cave.settlement, tell.settlement) %>%
    summarise(settlements = sum(settlement)/n()*100,
              open.settlement = sum(open.settlement)/n()*100,
              cave.settlement = sum(cave.settlement)/n()*100,
              tell.settlement = sum(tell.settlement)/n()*100
              )

```

**Excercise:**

1. Find other ways to answer the question.
2. Discuss the different advantages of your approach(es) and those presented above

***

Ok, let's continue. First, get rid of all the unnecessary variables. Again, there are many different possibilities to do this.

Here is a "classic" way:

```{r}
selected.data <- raw.data[,c(2,4:20,22:31,35:36)]
#knitr::kable(head(selected.data))
str(selected.data)
```

And here a more modern way:

```{r}
library(dplyr)
selected.data <- select(.data = raw.data,
                        -Nr.,
                        -starts_with("Site.ty"),
                        -starts_with("Ep"),
                        -starts_with("Code")
                        )
str(selected.data)
```

**Question:** Which way do you prefer? And besides your opinion, which is more suitable/robust [if any]?

## Create chronologically separated dataset ##

Now, we will create site data sets for different chronological periods. Again there are different ways to achieve this:

classic

```{r}
prehist <- selected.data[selected.data$prehistoric==1,]
table(selected.data$prehistoric)[2]==length(prehist$prehistoric)
```

modern

```{r}
library(dplyr)
prehist <- prehist %>%
    filter(prehistoric == 1) %>%
    select(-contains("roman")) %>%
    select(-contains("medie")) %>%
    select(-contains("modern"))

str(prehist)
table(selected.data$prehistoric)[2]==length(prehist$prehistoric)
```

**Question:** What does the line beginning with "table(..." mean?

Now that we have a clean dataset, let's plot some of its characteristics. As always we are going to use the standard and a modern approach.


The standard plotting way.

```{r}
barplot(cbind(paleolithic = sum(prehist$palaeolithic),
              neolithic = sum(prehist$neolithic),
              copper_age = sum(prehist$copper.age),
              bronze_age = sum(prehist$bronze.age),
              iron_age = sum(prehist$iron.age)
              ),
        main = "Number of sites from prehistoric periods"
        )
```

And so forth...

**Exercise:**

1. Plot the different numbers of site types.
2. Which way do you chose? Why? What other possibilities are there?

For the modern way we need to transform the data first from the a wide to a long format. Infos can be found e.g., [here](http://seananderson.ca/2013/10/19/reshape.html) or @wickham2007. In the case of wide data there is a column for each variable. 

Let's first take care of the data structure. We want to investigate the amount of different sity types per period.

```{r}
types.per.period <- data.frame(palaeolithic = 0,
                                     neolithic = 0,
                                     copper.age = 0,
                                     bronze.age = 0,
                                     iron.age = 0
                                     )

names(prehist)

library(magrittr); library(dplyr)
for (i in 3:18) {
    j <- i-2
    types.per.period[j,] <- prehist %>%
        filter(.[i] == 1) %>%
        summarise(paleolithic = sum(palaeolithic)/length(.),
                  neolithic = sum(neolithic),
                  copper.age = sum(copper.age),
                  bronze.age = sum(bronze.age),
                  iron.age = sum(iron.age)
                  )
}

str(types.per.period)
types.per.period$type <- names(prehist)[3:18]
types.per.period
```

This was again just data crunching/selection. Now, we change the data from wide to long format using the `reshape2` package.

```{r}
library(reshape2)
types2 <- melt(types.per.period, id.vars = c("type"))
str(types2)
head(types2)
```

**Question:** Do you see the difference? Describe the data new structure. 

From this point on it is very easy and effortless to produce informative plots. We use the `ggplot2` package for this. Examples and documentation can be found at http://docs.ggplot2.org/current/ or in the [manual](https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf) or in @wickham2009 as well as @zuur2009 (for the last on check this [link](http://www.cookbook-r.com/Graphs/)). `ggplot2` package is build on the ideas of "Grammar of graphics". More information about this broad topic can be found HERE and in @wilkinson2005.

```{r}
library(ggplot2)
ggplot(types2, aes(x=type,y=value, fill=variable)) +
    geom_bar(stat="identity")

```

**Exercises**

1. Check the help of ggplot and reference pages on the internet to:
   - change the theme of the plot; Don't know what a "theme" is? Check help!
   - reorder the x-axis.
   - give a title
   - change the direction of x-labels to make them readable
2. Create a facet plot differentiated by period.

In the end it might look like this and this


```{r, echo = FALSE}
## https://rstudio-pubs-static.s3.amazonaws.com/7433_4537ea5073dc4162950abb715f513469.html
types2$type <- factor(types2$type, levels = names(prehist)[3:18])

library(ggplot2)
ggplot(types2, aes(x=type,y=value, fill=variable)) +
    geom_bar(stat="identity") +
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(list(title = "Site types per period",
              x = "Type of archaeological site",
              y = "Absolute number of finds")
         )

ggplot(types2, aes(x=type,y=value/sum(value), fill=variable)) +
    geom_bar(stat="identity") +
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(list(title = "Site types (relative amount) per period",
              x = "Type of archaeological site",
              y = "Relative number of finds")
         )


ggplot(types2, aes(x=type,y=value)) +
    geom_bar(stat="identity") +
    facet_grid(variable~.) +
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90),
          strip.text.x = element_text(size = 10, angle = 0)
          ) +
    labs(list(title = "Site types per period",
              x = "Type of archaeological site",
              y = "Absolute number of finds")
         )

ggplot(types2, aes(x=variable,y=value, fill=type)) +
    geom_bar(stat="identity") + 
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(list(title = "Site types per period",
              y = "Relative amount of types of archaeological sites",
              x = "Period")
         )

ggplot(types2, aes(x=variable,y=value, fill=type)) +
    geom_bar(stat="identity", position = "fill") +
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(list(title = "Site types per period",
              y = "Relative amount of types of archaeological sites",
              x = "Period")
         )

```



***

# References #
