---
title: Inspect data and ask questions
author: Freie Universität Berlin -- M.Sc. Landschaftsarchäologie <br> Modellieren in der Landschaftsarchäologie <br> Dr. Daniel Knitter (CAU Kiel) <br>
date: WS 2016/2017
email: daniel.knitter@fu-berlin.de
bibliography: VS_Modellierung_Literature.bib
csl: harvard1.csl
output:
  ioslides_presentation:
    toc: true
    highlight: pygments
    widescreen: true
    smaller: true	
---

# Inspect data and ask questions

# Prerequisites #

## Set working directory ## {.build}

```{r echo=FALSE}
## this is here, so that knitr allows to change the working directory
knitr::opts_knit$set(root.dir = "..")
```

```{r eval=FALSE}
setwd("YOUR-PATH")
```	

The working directory should be at the top of your project directory structure. 

## Load data ## {.build}

Get the data from xls(x) files -- load required package [always troublesome on windows due to missing perl]

Steps:

- get help
- load required package

```{r message=FALSE, eval=FALSE}
##??xls
library(gdata)
raw_data <- read.xls(xls = "./data/Fundstellen_UTM.xlsx", sheet = 1)
```

A nice tutorial on loading table (including xls) data to `R` is [on datacamp](https://www.datacamp.com/community/tutorials/r-tutorial-read-excel-into-r#gs.5cgO=As)

## Load data ## {.build}

A better way: use text files. 

**Question: Why?** 

```{r load-data}
raw_data <- read.table(file = "data/FS_UTM.csv",
                       header = TRUE,
                       sep = ";",
                       stringsAsFactors = FALSE,
                       quote = "")
```

# Play with the data #

## Check structure and content ## {.build}

```{r}
str(raw_data)
```

## Ask questions ## {.build}

How many archaeological sites are in the data set?
   
```{r}
length(raw_data$Nr.)
```

**Exercise:** Show another way how to answer the question?

## Ask questions ## {.build}

How many settlements?

```{r}
sum(raw_data$settlement)
```

**Question:**

1. What other possibilities are there to query the amount of settlements?
2. What other question could be asked?


# Create a dataset for further analyses #

## Clean data ## {.build}

**Some mandatory things: **

- Are there missing values? 
- Do you understand every column/variable? 
- Where are problems with the data? 
- Typos? 
- What data are necessary/unnecessary?

The easiest but sometimes quite verbose way to check:

```{r}
str(raw_data)
```

## Check data structure 2 ##

Examples of more controlled queries:

```{r}
head(names(raw_data))
head(sapply(raw_data, class))
```

## Check data structure 2a ##

Examples of more controlled queries (`head`, `tail`, `sample`):

```{r}
sample(raw_data$Epoch..site., 3)
```

The structure of the "Epoch..site." column is a little complicated to
use for further analyses. Fortunately, we do not need to use them,
since there are already separated and precoded into binary variables.

**Questions: **

1. What is the problem with column?
2. What does "binary variable" mean? 
3. Why is it useful? 
4. Are there other ways to code a variable?

## Check data structure 2a ##

**Exercise [for the rest of the course-week...] **

Clean the messed variables of the table yourself!

- Hint 1: try `?gsub`
- Hint 2:

```{r, eval=FALSE}
grep(pattern = "Bronze Age", x = raw_data$Epoch..site.)
```

## Further questions 1 ##  {.build}

**What is the proportion of the different settlement types?**

One ordinary way (by the way, giving variable names "a","b", etc. is really **bad** style!)

```{r}
a <- raw_data[,4:7]
b <- cbind(sum(a[,1]),sum(a[,2]),sum(a[,3]),sum(a[,4]))
colnames(b) <- c("settlement","open.settlement","cave.settlement","tell.settlement")
b
c <- b/b[1]*100
c
```

## Further questions 2 ##

**What is the proportion of the different settlement types?**

One of the *apply ways

```{r}
a <- data.frame(settlement = raw_data$settlement[raw_data$settlement==1],
                open.settlement = raw_data$open.settlement[raw_data$settlement==1],
                cave.settlement = raw_data$cave.settlement[raw_data$settlement==1],
                tell.settlement = raw_data$tell.settlement[raw_data$settlement==1]
                )
sapply(a, sum)
sapply(a, function(x){sum(x)/length(x)*100})
```

## Further questions 3 ##

**What is the proportion of the different settlement types?**

The [magrittr](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) and [dplyr](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) way (Version 1):

```{r message=FALSE}
library(magrittr)
library(dplyr)

raw_data %>%
    filter(settlement==1) %>%
    select(settlement, open.settlement, cave.settlement, tell.settlement) %>%
    summarise(settlement=sum(.[,1])/n()*100,
              open.settlement=sum(.[,2])/n()*100,
              cave.settlement=sum(.[,3])/n()*100,
              tell.settlement=sum(.[,4])/n()*100
              )
```

## Further questions 3a ##

**What is the proportion of the different settlement types?**

The [magrittr](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) and [dplyr](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) way (Version 2):

```{r message=FALSE}
library(magrittr)
library(dplyr)

raw_data %>%
    filter(settlement==1) %>%
    select(settlement, open.settlement, cave.settlement, tell.settlement) %>%
    summarise(settlements = sum(settlement),
              open.settlement = sum(open.settlement),
              cave.settlement = sum(cave.settlement),
              tell.settlement = sum(tell.settlement)
              )
```

## Further questions 3b ##

**What is the proportion of the different settlement types?**

The [magrittr](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) and [dplyr](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) way (Version 2a):

```{r message=FALSE}
library(magrittr)
library(dplyr)
raw_data %>%
    filter(settlement==1) %>%
    select(settlement, open.settlement, cave.settlement, tell.settlement) %>%
    summarise(settlements = sum(settlement)/n()*100,
              open.settlement = sum(open.settlement)/n()*100,
              cave.settlement = sum(cave.settlement)/n()*100,
              tell.settlement = sum(tell.settlement)/n()*100
              )
```

**Excercise:**

1. Find other ways to answer the question.
2. Discuss the different advantages of your approach(es) and those presented above.

## Further questions 4 ##

What the heck...!? Why not using `table` and `prop.table` ?!

**Excercise:** Why? :)

## Clean data continued ##

Ok, let's continue. First, get rid of all the unnecessary variables. Again, there are many different possibilities to do this.

Here is a "classic" way (approach: I want to have...):

```{r}
selected_data <- raw_data[,c(2,4:20,22:31,35:36)]
names(selected_data)
#knitr::kable(head(selected_data))
#str(selected_data)
```

## Clean data continued ##

Ok, let's continue. First, get rid of all the unnecessary variables. Again, there are many different possibilities to do this.

Here is a "classic" way (approach: I __do not__ want to have...):

```{r}
selected_data <- raw_data[,c(-1,-3,-21,-32:-34)]
names(selected_data)
#knitr::kable(head(selected_data))
#str(selected_data)
```

## Clean data continued ##

Ok, let's continue. First, get rid of all the unnecessary variables. Again, there are many different possibilities to do this.

And here a more modern way (Question: Which approach is used?):

```{r}
library(dplyr)
selected_data <- select(.data = raw_data,
                        -Nr.,
                        -starts_with("Site.ty"),
                        -starts_with("Ep"),
                        -starts_with("Code")
                        )
names(selected_data)
#str(selected_data)
```

**Question:** Which way do you prefer? And besides your opinion, which is more suitable/robust [if any]?

## Create chronologically separated dataset 1 ## 

Now, we will create site data sets for different chronological periods. Again there are different ways to achieve this:

classic

- 1. select relevant data
- 2. check whether everything is fine
- 3. drop unnecessary columns

## Create chronologically separated dataset 1 ## 

```{r}
prehist <- selected_data[selected_data$prehistoric==1,]

table(selected_data$prehistoric)[2]==length(prehist$prehistoric)

prehist <- prehist[,-c(19,25:28)]
#str(prehist)
```

**Question:** What does the line beginning with "table(..." mean? 

## Create chronologically separated dataset 2 ##

a modern way

```{r}
library(dplyr)
prehist <- selected_data %>%
    filter(prehistoric == 1) %>%
    select(-contains("preh")) %>%
    select(-contains("roman")) %>%
    select(-contains("medie")) %>%
    select(-contains("modern"))

#str(prehist)
table(selected_data$prehistoric)[2]==length(prehist$prehistoric)
```

**Question:** Why is the "table(..." line now FALSE?

## at last: something to see {.columns-2 .build}

Now that we have a clean dataset, let's plot some of its characteristics. As always we are going to use the standard and a modern approach.

The standard plotting way.

```{r fig.height=5, fig.width=5}
barplot(cbind(
    paleolithic = sum(prehist$palaeolithic),
    neolithic = sum(prehist$neolithic),
    copper_age = sum(prehist$copper.age),
    bronze_age = sum(prehist$bronze.age),
    iron_age = sum(prehist$iron.age)
),
main = "Number of sites from prehistoric periods"
)
```

## at last: something to see

**Exercise:**

1. Plot the different numbers of site types.
2. Which way do you chose? Why? What other possibilities are there?

```{r fig.height=4, fig.width=10, echo=FALSE}
barplot(cbind(
    paleolithic = sum(prehist$palaeolithic[prehist$cave.settlement==1]),
    neolithic = sum(prehist$neolithic[prehist$cave.settlement==1]),
    copper_age = sum(prehist$copper.age[prehist$cave.settlement==1]),
    bronze_age = sum(prehist$bronze.age[prehist$cave.settlement==1]),
    iron_age = sum(prehist$iron.age[prehist$cave.settlement==1])
),
main = "Number of cave.settlements in prehistoric periods"
)
```

## at last: something to see -- GGPLOT2

For the modern way we need to transform the data first from the a wide to a long format. Infos can be found e.g., [here](http://seananderson.ca/2013/10/19/reshape.html) or @wickham2007 or @wickham2014. In the case of wide data there is a column for each variable. 

Let's first take care of the data structure. We want to investigate the amount of different sity types per period.

```{r}
types.per.period <- data.frame(palaeolithic = 0,
                               neolithic = 0,
                               copper.age = 0,
                               bronze.age = 0,
                               iron.age = 0
                               )
names(prehist)
```

## at last: something to see -- GGPLOT2

Data crunching

```{r}
library(magrittr); library(dplyr)
for (i in 3:18) {
    j <- i-2
    types.per.period[j,] <- prehist %>%
        filter(.[i] == 1) %>%
        summarise(paleolithic = sum(palaeolithic),
                  neolithic = sum(neolithic),
                  copper.age = sum(copper.age),
                  bronze.age = sum(bronze.age),
                  iron.age = sum(iron.age)
                  )
}

#str(types.per.period)
types.per.period$type <- names(prehist)[3:18]
types.per.period
```

## at last: something to see -- GGPLOT2

This was again just data crunching/selection. Now, we change the data from wide to long format using the `reshape2` package.

```{r}
library(reshape2)
types2 <- melt(types.per.period, id.vars = c("type"))
str(types2)
head(types2, 3)
```

**Question:** Do you see the difference? Describe the data new structure. 

## at last: something to see -- GGPLOT2

From this point on it is very easy and effortless to produce informative plots. We use the `ggplot2` package for this. Examples and documentation can be found at http://docs.ggplot2.org/current/ or in the [manual](https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf) or in @wickham2009 as well as @zuur2009 (for the last on check this [link](http://www.cookbook-r.com/Graphs/)).

`ggplot2` package is build on the ideas of "Grammar of graphics". More information about this broad topic can be found in @wilkinson2005.

`ggplot2` expects "tidy" data. To get to know what this exactly is, look at @wickham2014.

## at last: something to see -- GGPLOT2

```{r}
library(ggplot2)
ggplot(types2, aes(x=type,y=value, fill=variable)) +
    geom_bar(stat="identity")

```

## at last: something to see -- GGPLOT2

**Exercises**

1. Check the help of ggplot and reference pages on the internet to:

	- change the theme of the plot; Don't know what a "theme" is? Check help!
	- reorder the x-axis.
	- give a title
	- change the direction of x-labels to make them readable
	
2. Create a facet plot differentiated by period.

In the end it might look like this and this...

## at last: something to see -- GGPLOT2

```{r, echo = FALSE}
## https://rstudio-pubs-static.s3.amazonaws.com/7433_4537ea5073dc4162950abb715f513469.html
types2$type <- factor(types2$type, levels = names(prehist)[3:18])

library(ggplot2)
ggplot(types2, aes(x=type,y=value, fill=variable)) +
    geom_bar(stat="identity") +
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(list(title = "Site types per period",
              x = "Type of archaeological site",
              y = "Absolute number of finds",
              fill = "Period")
         )
```

## at last: something to see -- GGPLOT2

```{r, echo = FALSE}   
ggplot(types2, aes(x=type,y=value/sum(value), fill=variable)) +
    geom_bar(stat="identity") +
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(list(title = "Site types (relative amount) per period",
              x = "Type of archaeological site",
              y = "Relative number of finds",
              fill = "Period")
         )
```

## at last: something to see -- GGPLOT2

```{r, echo = FALSE}   
ggplot(types2, aes(x=type,y=value)) +
    geom_bar(stat="identity") +
    facet_grid(variable~.) +
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90),
          strip.text.x = element_text(size = 10, angle = 0)
          ) +
    labs(list(title = "Site types per period",
              x = "Type of archaeological site",
              y = "Absolute number of finds")
         )
```

## at last: something to see -- GGPLOT2

```{r, echo = FALSE}   
ggplot(types2, aes(x=variable,y=value, fill=type)) +
    geom_bar(stat="identity") + 
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(list(title = "Site types per period",
              y = "# of types of archaeological sites",
              x = "Period")
         )
```

## at last: something to see -- GGPLOT2

```{r, echo = FALSE}   
ggplot(types2, aes(x=variable,y=value, fill=type)) +
    geom_bar(stat="identity", position = "fill") +
    theme_bw(base_family = "Open Sans Light") +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(list(title = "Site types per period",
              y = "Relative # of types of archaeological sites",
              x = "Period")
         )

```

# References #

## 
