---
title: "Spatial Data"
author: Freie Universität Berlin -- M.Sc. Landschaftsarchäologie <br> Modellieren in der Landschaftsarchäologie <br> Dr. Daniel Knitter (CAU Kiel) <br>
date: WS 2016/2017
bibliography: VS_Modellierung_Literature.bib
csl: harvard1.csl
output:
  ioslides_presentation:
    toc: true
    highlight: pygments
    widescreen: true
    smaller: true
---

# Prerequisites

## setwd, library, data {.build}

```{r echo=FALSE}
## this is here, so that knitr allows to change the working directory
knitr::opts_knit$set(root.dir = "..")
```

```{r eval=FALSE}
setwd("YOUR-PATH")
```	

```{r load-data}
raw_data <- read.table(file = "data/FS_UTM.csv",
                       header = TRUE,
                       sep = ";",
                       stringsAsFactors = FALSE,
                       quote = "")
```

## 4a. Check data structure ## {.build}

Do you remember the "4" issue from the previous session?! Let's check it again and solve the problem. I use a `(s)apply` way. You can chose the method you prefer.

First, let's check which columns have problems

```{r message=FALSE}
#str(raw_data)
#sapply(raw_data[,4:20], function(x){x[x>1]})
apply(X=raw_data[,4:20], MARGIN = 2, FUN=function(x){x[x>1]})
```

## 4a. Check data structure ## {.build}

Now, let's solve the problem and check again

```{r message=FALSE}
raw_data$settlement[raw_data$settlement > 1] <- 1

sapply(raw_data[,4:20], function(x){x[x>1]})
```

## 4b. Check data structure ## {.build}

So, now it's time to have a look at the coordinates...

```{r message=FALSE}
#str(raw_data)
raw_data[c(1:6),c(2,35,36)]
```

**Question:** Is there another way to create this overview?

## 4b. Check data structure

Yes.

```{r message=FALSE}
#str(raw_data)
head(raw_data[,c(2,35,36)])
```

# Create a spatial object #

## Some useful references ##

> - THE book that covers a broad range of topics:
    + Bivand, Roger S., Edzer J. Pebesma, and Virgilio Gómez-Rubio. 2008. Applied Spatial Data Analysis with R. New York: Springer.
> - Useful tutorial-like introduction in the handling and plotting of spatial data
    + Lovelace, Robin, and James Cheshire. 2015. “Introduction to Visualising Spatial Data in R.” [link](https://github.com/Robinlovelace/Creating-maps-in-R/raw/master/intro-spatial-rl.pdf).

## Spatial data libraries ## 

```{r message=FALSE}
library(sp)
library(rgdal)
```

> - sp: Classes and Methods for Spatial Data
    + https://cran.r-project.org/web/packages/sp/index.html
> - rgdal: Bindings for the Geospatial Data Abstraction Library
    + based on GDAL "Swiss Army Knife for Geospatial Data" -> https://gdal.org
    + https://cran.r-project.org/web/packages/rgdal/index.html


## Create spatial object ##

As always there are numerous ways to accomplish this. See https://cran.r-project.org/web/packages/sp/vignettes/intro_sp.pdf

There are some things we need to know in advance:

> 1. What kind of spatial data do we have?
> 2. What are the coordinates?
> 3. What is the reference system?
    + Your best friend: https://www.epsg.io

## Create spatial data ##

```{r}
sites <- SpatialPointsDataFrame(coords = cbind(raw_data$xUTM,raw_data$yUTM),
                                data = data.frame(raw_data[,2]),
                                proj4string = CRS("+init=epsg:32634")
                                )
class(sites)
```

**Exercise:** Find other ways to create a spatial data object.

## Create spatial data ## {.build}

```{r}
sites <- raw_data
coordinates(sites) <- ~xUTM+yUTM
class(sites)
```

**Exercise:** Check the `str`ucture of the `sites` object. What is missing?

```{r}
is.projected(sites)
```

```{r}
proj4string(sites) <- CRS("+init=epsg:32634")
is.projected(sites)
```

## Plot it

```{r fig.height=3, fig.width=5}
plot(sites)
```

Find further ways to plot the points.

A collection of examples: https://pakillo.github.io/R-GIS-tutorial/

## Plot it 2.0  {.columns-2 .build}

It's all about interaction, isn't it?

```{r fig.height=5, fig.width=5}
library(magrittr)
library(leaflet)

sites <- spTransform(sites, "+init=epsg:4326")

m1 <- leaflet() %>%
    addProviderTiles("Thunderforest.Landscape") %>%
        addMarkers(lng=sites@coords[,1],
                   lat=sites@coords[,2],
                   popup = sites@data$Site.Name
                   )
m1
```

## Plot it 2.0  {.columns-2 .build}

You can also change the map tile provider for a wide range of different maps. An overview can be found here: http://leaflet-extras.github.io/leaflet-providers/preview/index.html

```{r fig.height=5, fig.width=5}
library(magrittr)
library(leaflet)

sites <- spTransform(sites, "+init=epsg:4326")

m2 <- leaflet() %>%
    addProviderTiles("OpenStreetMap.HOT") %>%
        addMarkers(lng=sites@coords[,1],
                   lat=sites@coords[,2],
                   popup = sites@data$Site.Name
                   )
m2
```

## Plot it 2.0

Another great package for interactive visualization: [mapview](http://environmentalinformatics-marburg.github.io/web-presentations/20150723_mapView.html)

**Question:** What is the drawback of these interactive mapping approaches?

## Export to Shapefile

```{r eval=TRUE}
writeOGR(obj = sites,
         dsn = "data",
         layer = "A_Sites",
         driver = "ESRI Shapefile",
         overwrite_layer = TRUE
         )

list.files("./data")
```

## Export to graphics file

not very useful; just to show how easy it is

```{r, eval=FALSE}
png("data/A_Sites.png")
plot(sites)
dev.off()

## pdf, svg, postscript
?grDevices

?cairo
```

# Play. 


