---
title: "Raster analysis"
author: Freie Universität Berlin -- M.Sc. Landschaftsarchäologie <br> Modellieren in der Landschaftsarchäologie <br> Dr. Daniel Knitter (CAU Kiel) <br>
date: WS 2016/2017
bibliography: VS_Modellierung_Literature.bib
csl: harvard1.csl
output:
  ioslides_presentation:
    toc: true
    highlight: pygments
    widescreen: true
    smaller: true
---

# Prerequisites

## setwd, library, data, cleaning

```{r echo=FALSE}
## this is here, so that knitr allows to change the working directory
knitr::opts_knit$set(root.dir = "..")
```

```{r eval=FALSE}
setwd("YOUR-PATH")
```	

```{r load-data, message=FALSE}
raw_data <- read.table(file = "data/FS_UTM.csv",
                       header = TRUE,
                       sep = ";",
                       stringsAsFactors = FALSE,
                       quote = "")

## check for issues/problems
## str(raw_data)
## apply(X=raw_data[,4:20], MARGIN = 2, FUN=function(x){x[x>1]})
raw_data$settlement[raw_data$settlement > 1] <- 1

library(sp)
sites <- raw_data
coordinates(sites) <- ~xUTM+yUTM
proj4string(sites) <- CRS("+init=epsg:32634")
```

# Get data #

## SRTM scenes and rgdal

Please go to http://srtm.csi.cgiar.org/, download the appropriate SRTM scence(s) for our Romanian research area, unzip them and load them into `R` using the `rgdal` package.

Plot the sites onto the raster to see whether the srtm scene covers all the points.

```{r message=FALSE}
library(rgdal)
srtm <- readGDAL("data/srtm_41_03.tif")
```

## SRTM scenes and rgdal

```{r message=FALSE}
image(srtm)
points(spTransform(sites, CRS("+init=epsg:4326")))
```

## SRTM scences and the raster package

```{r}
library(raster)
srtm <- raster("data/srtm_41_03.tif")
plot(srtm)
points(spTransform(sites, CRS("+init=epsg:4326")))
```

## Reprojecting

```{r}
## takes a bit of time
srtm <- projectRaster(srtm, res=90, crs=CRS("+init=epsg:32634"))
plot(srtm)
points(sites)
```

## Cropping

```{r}
srtm <- crop(x = srtm, y = extent(sites)+20000)
plot(srtm)
points(sites)
```

## Terrain analyses

Using the `raster` package we can create a lot of different terrain parameters by just one line of code. The result will be a multi-layer raster object, like you know it from multi-/hyperspectral satellite images. You can work on such scenes like you would work on ordinary vector objects in `R`.

```{r terrain_parameter, echo=TRUE}
srtm.tp <- terrain(x = srtm,
                   opt = c("slope",
                           "aspect",
                           "TPI",
                           "TRI",
                           "roughness",
                           "flowdir"),
                   unit = "degrees",
                   neighbors = 8)
```

## Terrain analyses

```{r echo=TRUE}
plot(srtm.tp)
```

## Terrain analyses

You should be familiar with slope and aspect but do you know the what the other parameter are? A look in the help offers insights (they are based on @wilson2000):

- "TRI (Terrain Ruggedness Index) is the mean of the absolute differences between the value of a cell and the value of its 8 surrounding cells.
- TPI (Topographic Position Index) is the difference between the value of a cell and the mean value of its 8 surrounding cells.
- Roughness is the difference between the maximum and the minimum value of a cell and its 8 surrounding cells" (from `?raster::terrain`)

The help also explains that we can use focal functions in order the adapt the approaches to our needs. A *focal function* corresponds to a moving window. It uses a matrix of weights for the neighborhood of the focal cells.

## TPI for different neighborhood size

```{r tpi_focal}
tpiw <- function(x, w=5) {
    m <- matrix(1/(w^2-1), nc=w, nr=w) # mean of the surrounding pixel
    m[ceiling(0.5 * length(m))] <- 0 # set the centre cell 0
    f <- focal(x, m, pad = TRUE) # apply moving window
    x - f
}
tpi15 <- tpiw(x = srtm, w=15)
tpi31 <- tpiw(x = srtm, w=31)
#par(mfrow=c(1,3))
#plot(srtm.tp$tpi)
#plot(tpi15)
#plot(tpi31)
```

## TPI for different neighborhood size

For better visualisation we calculate a hillshade and overlay the TPI raster.

```{r tpi-plot, echo=TRUE}
srtm.hs <- hillShade(slope = terrain(srtm, opt="slope"),
                     aspect = terrain(srtm, opt="aspect"),
                     angle= 150, direction = 45, normalize = TRUE)
```

## TPI for different neighborhood size

```{r}
plot(srtm.hs,
     col = gray.colors(n= 255, start=.2, end=.9),
     legend = FALSE)
plot(tpi31,
     col=colorRampPalette(c("red", "white", "blue"))(255),
     alpha = .5, add=TRUE)
```

## Export raster

```{r}
writeRaster(x = tpi31,
            filename = "./results/tpi31.tif",
            overwrite = TRUE
            )
list.files(path = "./results")
```

## Extract values {.columns-2}

Now, it is time to integrate raster and points.

```{r fig.height=5,fig.width=5}
tmp <- extract(x = tpi31, y = sites)
head(tmp)
length(tmp)
hist(tmp)
```

## another way to plot

```{r tpi-plot2, echo=TRUE, message=FALSE}
library(rasterVis)
#srtmTheme <- rasterTheme(region=terrain.colors(200))
#levelplot(srtm, par.settings = srtmTheme)
levelplot(brick(tpi31, tpi15,srtm.tp$tpi), par.settings = RdBuTheme)
```


# References

## 
