---
title: "Interpolation"
author: Freie Universität Berlin -- M.Sc. Landschaftsarchäologie <br> Modellieren in der Landschaftsarchäologie <br> Dr. Daniel Knitter (CAU Kiel) <br>
date: WS 2016/2017
bibliography: VS_Modellierung_Literature.bib
csl: harvard1.csl
output:
  ioslides_presentation:
    toc: true
    highlight: pygments
    widescreen: true
    smaller: true
---

# Prerequisites

## setwd, library, data, cleaning

```{r echo=FALSE}
## this is here, so that knitr allows to change the working directory
knitr::opts_knit$set(root.dir = "..")
```

```{r eval=FALSE}
setwd("YOUR-PATH")
```	

```{r load-data, message=FALSE}
raw_data <- read.table(file = "data/FS_UTM.csv",
                       header = TRUE,
                       sep = ";",
                       stringsAsFactors = FALSE,
                       quote = "")

## check for issues/problems
## str(raw_data)
## apply(X=raw_data[,4:20], MARGIN = 2, FUN=function(x){x[x>1]})
raw_data$settlement[raw_data$settlement > 1] <- 1

library(sp)
sites <- raw_data[(raw_data$settlement == 1 & raw_data$iron.age == 1),]
coordinates(sites) <- ~xUTM+yUTM
proj4string(sites) <- CRS("+init=epsg:32634")
```

## Create Point Pattern dataset

```{r message=FALSE}
library(spatstat)
                                                    
sites_pp <- ppp(x = sites@coords[,1],
                y = sites@coords[,2],
                window = owin(xrange = c(min(sites@coords[,1]),
                                         max(sites@coords[,1])
                                         ),
                              yrange = c(min(sites@coords[,2]),
                                         max(sites@coords[,2])
                                         )
                              )
                )
unitname(sites_pp) <- c("meter", "meters")

anyDuplicated(sites_pp)
sites_pp <- sites_pp[duplicated(sites_pp)==FALSE]
anyDuplicated(sites_pp)
#plot(sites_pp)
```

## Caluclate density using graph methods

Distance based density using "largest empty circle"

Idea: calculate the centres of the largest empty circles -- corresponds to the edges of the Voronoi graph

```{r}
library(tripack)

sites_vor <- voronoi.mosaic(x = sites_pp$x,
                            y = sites_pp$y,
                            duplicate = "remove"
                            ) 
rad <- sites_vor$radius
lec <- SpatialPointsDataFrame(coords = cbind(sites_vor$x, sites_vor$y),
                              data = as.data.frame(rad), # length of edge
                              proj4string=CRS("+init=epsg:32634")
                              )
```

## Caluclate density using graph methods {.columns-2}

```{r, fig.height=4, fig.width=4}
plot(lec)
## remove outlier
lec <- lec[lec@coords[,2]>min(lec@coords[,2]),]
plot(lec)
```

## Caluclate density using graph methods

```{r}
plot(sites_pp)
plot(sites_vor, col = "red", add=TRUE)
```

## Caluclate density using graph methods {.columns-2}

```{r message=FALSE, fig.height=5, fig.width=5}
plot(sites_pp, pch = 19)
points(lec, col="red")

library(plotrix)
for (i in 1:length(lec@coords[,1])) {
    draw.circle(x = lec@coords[i,1],
                y = lec@coords[i,2],
                radius = lec@data$rad[i],
                border = "gray")
    }
```

## Interpolation

We're lazy...instead of creating a raster we use one that is "already there":

```{r message=FALSE}
library(raster)
srtm <- raster("results/srtm.tif")
srtm <- aggregate(x = srtm, fact = 9)
library(maptools)
srtm <- as(object = srtm,
           Class = "SpatialGridDataFrame"
           )
str(srtm)
##image(srtm)
##points(sites, pch=19, cex=.5)
```

## Interpolation - Inverse Distance Weighting

```{r message=FALSE}
library(gstat)
crs(srtm) <- lec@proj4string ## BE CAREFUL!

lec_gstat_p2 <- gstat(formula = lec@data$rad ~ 1,
                      locations = lec,
                      set = list(idp = 2)
                      )
lec_gstat_p25 <- gstat(formula = lec@data$rad ~ 1,
                       locations = lec,
                       set = list(idp = 2.5)
                       )
lec_idw2 <- predict(object = lec_gstat_p2,
                    newdata = srtm)
head(as.data.frame(lec_idw2))
lec_idw25 <- predict(object = lec_gstat_p25,
                     newdata = srtm)
```

## Interpolation - Inverse Distance Weighting {.columns-2}

```{r fig.height=5, fig.width=5, echo=FALSE}
plot(raster(lec_idw25))
points(sites_pp, col = rgb(0,0,0,.3)) 

plot(raster(lec_idw2))
points(sites_pp, col = rgb(0,0,0,.3)) 
```
 
## Interpolation - Trend surfaces {.columns-2}

```{r trendsurf, fig.width = 5, fig.height = 5}
lecp2 <- lm(
    lec@data$rad ~ poly(x = lec@coords[,1],
                        y = lec@coords[,2],
                        degree = 2
                        )
)
lecp3 <- lm(
    lec@data$rad ~ poly(x = lec@coords[,1],
                        y = lec@coords[,2],
                        degree = 3
                        )
)
plot(lec@data$rad ~ 1)
lines(lecp2$fitted.values, col = "red")
lines(lecp3$fitted.values, col = "blue")
legend("top",
       lty = c(1,1,1),
       col = c("red","blue","green"),
       legend = c("2nd degree polynom",
                  "3rd degree polynom"
                  )
       )
```

## Interpolation - Trend surfaces {.columns-2}

```{r trendsurf2, fig.width = 5, fig.height = 5}
lec_poly <- krige(
    formula = lec@data$rad ~ 1,
    locations = lec,
    newdata = srtm,
    degree = 3
)
plot(raster(lec_poly))
```

## Interpolation  {.columns-2}

Correlation between point pairs at different separation distances.

```{r hscat, message=FALSE, fig.width = 5, fig.height = 5}
library(lattice)
hscat(formula = lec@data$rad ~ 1,
      data = lec,
      breaks = seq(0, 10, 1)*1000
      )
```

## Interpolation -- Variogram

```{r vario1, fig.width = 10, fig.height = 4}
library(gstat)
plot(variogram(object = lec@data$rad ~ 1,
               locations = lec,
               cloud = TRUE
          )
     )
```

## Interpolation - Kriging  {.columns-2}

```{r, fig.width=4, fig.height=4}
library(gstat)

plot(variogram(lec$rad ~ 1, lec, cloud = TRUE))
```

```{r, fig.width=4, fig.height=4}
library(gstat)

plot(variogram(lec$rad ~ 1, lec))
```

## Interpolation -- Variogram {.columns-2}

```{r vario_rand, fig.height=5, fig.width=5}
library(lattice); library(gstat)

v <- variogram(lec@data$rad ~ 1, lec)
xyplot(
    x = gamma/1e8 ~ dist,
    data = v,
    pch = 3,
    type = 'b',
    lwd = 2,
    panel = function(x, y, ...) {
        for (i in 1:500) {
            lec$random = sample(lec$rad)
            v = variogram(random ~ 1, lec)
            llines(x = v$dist,
                   y = v$gamma/1e8,
                   col = rgb(.5,.5,.5,.2)
                   )
        }
        panel.xyplot(x, y, ...)
    },
    xlab = 'distance',
    ylab = 'semivariance/1e8'
)
```

## Interpolation 

```{r vario_vgms, fig.height=5, fig.width=10}
show.vgms()
```

## Interpolation {.columns-2}

```{r vario2, fig.height=5, fig.width=5}
plot(variogram(
    object = lec@data$rad ~ 1,
    locations = lec,
    boundaries = c(seq(0, 1e+3, 1e+2),
                   seq(2e+3, 1e+4, 1e+3)
                   ),
    cutoff = 40000
)
)
```

## Interpolation {.columns-2}

```{r vario2b, fig.height=5, fig.width=5}
plot(variogram(object = lec@data$rad ~ 1,
               locations = lec,
               cloud = FALSE,
               width = 1000,
               cutoff = 40000
          )
     )
```

## Interpolation - Kriging

```{r, fig.height=4, fig.width=10}
library(gstat)

plot(variogram(lec$rad ~ 1, lec, cutoff = 40000, alpha = c(0,45,90,135)))
```

## Interpolation - Kriging  {.columns-2}

When fitting goes wrong...

```{r, fig.height=5, fig.width=5}
vt <- variogram(object = lec@data$rad ~ 1,
               locations = lec,
               width = 5000,
               cutoff = 40000
          )
vt_vgm <- vgm(psill = 8.0e+07,
              model = "Gau",
              range = 40000,
              nugget = .2e+07)
v_fit <- fit.variogram(vt,vt_vgm)
plot(vt,v_fit)
```

## Interpolation -- tool: eyefit

```{r, eval=FALSE}
library(geoR)
v_eye <- eyefit(variog(as.geodata(lec["rad"]), max.dist = 40000))
ve_fit <- as.vgm.variomodel(v_eye[[1]])
v_fit <- fit.variogram(vt,ve_fit)
plot(vt,v_fit)
```

## Interpolation - Kriging  {.columns-2}

As it gets obvious: the chosen method (and/or) its application leads to completely useless results.

```{r, fig.height=5, fig.width=5}
lec_kri <- krige(formula = lec$rad ~ 1,
                 location = lec,
                 newdata = srtm,
                 model = v_fit
                 )

plot(raster(lec_kri),
     main="Kriging using Gaussian fit"
     )
points(sites,pch=20,cex=.4)
```

## Inverse Distance Weighting vs. Kriging {.columns-2}

```{r fig.height=5, fig.width=5, echo=FALSE}
plot(raster(lec_idw2), legend = FALSE)
points(sites_pp, cex = .5)
points(lec, pch = 19, cex = .4)
```

```{r fig.height=5, fig.width=5, echo=FALSE}
plot(raster(lec_kri), legend = FALSE)
points(sites_pp, cex = .5)
points(lec, pch = 19, cex = .4)
```

## Interpolation

So, did any of the things calculated so far make any sense?

Let's assume there is some value in it, what does it mean (landscape) archaeologically?

Why is the analysis at the current stage problematic?

What might be the reasons for the failure?

