---
title: "Interpolation"
author:
  Modellieren in der Landschaftsarchäologie
  Daniel Knitter
date: 2016-02-11
bibliography: VS_Modellierung_Literature.bib
csl: harvard1.csl
output:
  ioslides_presentation:
    toc: true
    highlight: pygments
    widescreen: true
    smaller: true
---

# Prerequisites

## setwd, library, data, cleaning

```{r message=FALSE}
setwd("/media/daniel/homebay/Teachings/WS_2015-2016/VS_Modellieren_Landschaftsarchaeologie")

raw.data <-  read.table(file = "./data/Fundstellen_UTM.csv",
                        header = TRUE,
                        sep = ";",
                        dec = ".",
                        quote = "",
                        stringsAsFactors = FALSE
                        )

## check for issues/problems
## str(raw.data)
## apply(X=raw.data[,4:20], MARGIN = 2, FUN=function(x){x[x>1]})
raw.data$settlement[raw.data$settlement > 1] <- 1

library(sp)
sites <- raw.data[(raw.data$settlement == 1 & raw.data$iron.age == 1),]
coordinates(sites) <- ~xUTM+yUTM
proj4string(sites) <- CRS("+init=epsg:32634")
```

## Create Point Pattern dataset

```{r message=FALSE}
library(spatstat)
                                                    
sites.pp <- ppp(x = sites@coords[,1],
                y = sites@coords[,2],
                window = owin(xrange = c(min(sites@coords[,1]),
                                         max(sites@coords[,1])
                                         ),
                              yrange = c(min(sites@coords[,2]),
                                         max(sites@coords[,2])
                                         )
                              )
                )
unitname(sites.pp) <- c("meter", "meters")

anyDuplicated(sites.pp)
sites.pp <- sites.pp[duplicated(sites.pp)==FALSE]
anyDuplicated(sites.pp)
#plot(sites.pp)
```

## Caluclate density using graph methods

Distance based density using "largest empty circle"

Idea: calculate the centres of the largest empty circles -- corresponds to the edges of the Voronoi graph

```{r}
library(tripack)

sites.vor <- voronoi.mosaic(x = sites.pp$x,
                      y = sites.pp$y,
                      duplicate = "remove"
                      ) 
rad <- sites.vor$radius
lec <- SpatialPointsDataFrame(coords = cbind(sites.vor$x, sites.vor$y),
                              data = as.data.frame(rad), # length of edge
                              proj4string=CRS("+init=epsg:32634")
                              )
```

## Caluclate density using graph methods

```{r}
plot(sites.pp)
plot(sites.vor, col = "red", add=TRUE)
```

## Caluclate density using graph methods {.columns-2}

```{r message=FALSE, fig.height=5, fig.width=5}
plot(sites.pp, pch = 19)
points(lec, col="red")

library(plotrix)
draw.circle(x = lec@coords[,1],
            y = lec@coords[,2],
            radius = sites.vor$radius,
            border = "gray")
```

## Interpolation

We're lazy...instead of creating a raster we use one that is "already there":
load SRTM scene, reproject, crop, change to SpatialGridDataFrame

```{r message=FALSE}
library(raster)
srtm <- raster("./data/srtm_41_03.tif")
srtm <- projectRaster(srtm, res=90, crs=CRS("+init=epsg:32634"))
srtm <- crop(x = srtm, y = extent(sites)+20000)
library(maptools)
srtm <- as(object = srtm,
           Class = "SpatialGridDataFrame"
           )
str(srtm)
##image(srtm)
##points(sites, pch=19, cex=.5)
```

## Interpolation - Inverse Distance Weighting

```{r message=FALSE}
library(gstat)

lec.gstat.p2 <- gstat(formula = lec@data$rad ~ 1,
                   locations = lec,
                   set = list(idp = 2)
                   )
lec.gstat.p05 <- gstat(formula = lec@data$rad ~ 1,
                   locations = lec,
                   set = list(idp = .5)
                   )
lec.idw2 <- predict(object = lec.gstat.p2,
                    newdata = srtm)
lec.idw05 <- predict(object = lec.gstat.p05,
                    newdata = srtm)
```

## Interpolation - Inverse Distance Weighting {.columns-2}

```{r fig.height=5, fig.width=5, echo=FALSE}
plot(raster(lec.idw05))
points(sites.pp)

plot(raster(lec.idw2))
points(sites.pp) 
```

## Interpolation - Kriging

**very** superficial

```{r}
library(gstat)

plot(variogram(lec$rad ~ 1, lec, cloud = TRUE))
```

## Interpolation - Kriging

**very** superficial

```{r}
library(gstat)

plot(variogram(lec$rad ~ 1, lec))
```

## Interpolation - Kriging

**very** superficial

```{r}
library(gstat)

plot(variogram(lec$rad ~ 1, lec, alpha = c(0,45,90,135)))
```

## Interpolation - Kriging {.columns-2}

**very** superficial

```{r fig.height=5, fig.width=5, echo=FALSE}
#vt <- variogram(lec$rad ~ 1, lec)
plot(variogram(lec$rad ~ 1, lec))
show.vgms()
```

## Interpolation - Kriging

**very** superficial

```{r}
vt <- variogram(lec$rad ~ 1, lec)
v.fit <- fit.variogram(vt,
                       vgm(nugget = 3e+07,
                           model = "Gau",
                           psill = 3e+08, #355572513
                           range = 180000)
                       )
```

## Interpolation - Kriging

**very** superficial

```{r}
plot(vt,v.fit)
```

## Interpolation - Kriging

**very** superficial

```{r}
lec.kri <- krige(formula = lec$rad ~ 1,
                 location = lec,
                 newdata = srtm,
                 model = v.fit
                 )
```

## Interpolation - Kriging

**very** superficial

```{r}
plot(raster(lec.kri),
     main="Kriging using Gaussian fit"
     )
points(sites,pch=20,cex=.4)
```

## Interpolation - Inverse Distance Weighting vs. Kriging {.columns-2}

```{r fig.height=5, fig.width=5, echo=FALSE}
plot(raster(lec.idw05))
points(sites.pp)

plot(raster(lec.kri))
points(sites.pp) 
```
