---
title: "Distance analyses"
author:
- Freie Universität Berlin -- M.Sc. Landschaftsarchäologie
- Modellieren in der Landschaftsarchäologie
- Dr. Daniel Knitter (CAU Kiel)
date: WS 2016/17
bibliography: VS_Modellierung_Literature.bib
csl: harvard1.csl
output:
  ioslides_presentation:
    toc: true
    highlight: pygments
    widescreen: true
    smaller: true
---

# Prerequisites

## setwd, library, data, cleaning

```{r echo=FALSE}
## this is here, so that knitr allows to change the working directory
knitr::opts_knit$set(root.dir = "..")
```

```{r eval=FALSE}
setwd("YOUR-PATH")
```	

```{r load-data, message=FALSE}
raw_data <- read.table(file = "data/FS_UTM.csv",
                       header = TRUE,
                       sep = ";",
                       stringsAsFactors = FALSE,
                       quote = "")

## check for issues/problems
## str(raw_data)
## apply(X=raw_data[,4:20], MARGIN = 2, FUN=function(x){x[x>1]})
raw_data$settlement[raw_data$settlement > 1] <- 1

library(sp)
sites <- raw_data[(raw_data$settlement == 1 & raw_data$iron.age == 1),]
coordinates(sites) <- ~xUTM+yUTM
proj4string(sites) <- CRS("+init=epsg:32634")
```

## Distance analyses in geographic space {.columns-2}

- abstract concept: can be measured in meters, time, money, strangeness, ...
- influence interactions

```{r fig.width=5, echo=FALSE}
library(png)
library(grid)
img <- readPNG("assets/img/pic_daniel_Harris_Ullman_1945_p8.png")
grid.raster(img)
```

## Distance analyses in geographic space

1. straight/ direct

	- Euclidean distance
	- "As the crow flies"
	- "In a beeline"
	- geodesic distance
	- great-circle distance

2. not direct/ random
		
	- random walk
	- "Drunkard's Walk"


## Distance analyses in geographic space

$$d(i,j) = \sqrt{(x_i-x_j)^2 + (y_i-y_j)^2}$$

## Distance analyses in geographic space

Distance between points using Euclidean distance using package `rgeos`

```{r, message=FALSE}
library(rgeos)
ed <- gDistance(spgeom1 = sites,
                byid = TRUE)
str(ed)
dim(ed)
head(ed[1,])
```

Can you imagine the structure if the `ed` object?

## Distance analyses in geographic space

```{r}
min(ed)
```

```{r}
round(data.frame(min = min(ed[ed>0]),
                 max = max(ed[ed>0]),
                 mean = mean(ed[ed>0]),
                 median = median(ed[ed>0])
                 ),
      2)
```

## Distance analyses in geographic space

```{r}
hist(ed)
abline(v = mean(ed[ed>0]), col = "red")
abline(v = median(ed[ed>0]), col = "blue")
```

## Distance analyses in geographic space

Some further questions that might be interesting

```{r}
mp <- apply(X = ed,
            MARGIN = 2,
            FUN = mean)

tmp <- ed
tmp[tmp==0] <- 9999

cp <- apply(X = tmp,
            MARGIN = 2,
            FUN = min)

sites$mp <- mp
sites@data$cp <- cp
```

What is the mean euclidean distance from a point to its surrounding points (mp)?

What is the closest euclidean distance from a point to its surrounding points (cp)?

**Question:** Why changing 0 to 9999?

## Distance analyses in geographic space

```{r fig.show="hide"}
spplot(sites, "cp")
spplot(sites, "mp")
```

Examples how to use `spplot`: http://rspatial.r-forge.r-project.org/gallery/

## Distance analyses in geographic space

Leaflet in `R`: https://rstudio.github.io/leaflet/

```{r eval=FALSE}
library(leaflet)
tmp <- spTransform(sites, CRSobj = CRS("+init=epsg:4326"))
leaflet(data=tmp$cp) %>%
    addTiles() %>%
    addCircleMarkers(lng = tmp@coords[,1],
                     lat = tmp@coords[,2],
                     radius = tmp$mp/1000,
                     )
```

## Distance analyses in geographic space

`Mapview` [as alternative to `leaflet`]: http://environmentalinformatics-marburg.github.io/mapview/introduction.html

```{r, message = FALSE}
library(mapview)
mapview(sites, zcol = "mp", legend = TRUE)
```

## Distance analyses in geographic space {.build}

**Least-cost distances**

- Cost is --- like distance --- an abstract concept; hence, YOU decide what your costs are (time, energy, ugliness, ...)
- Least-cost distance analyses allow to integrate prior knowledge and more complexity into your model.

**There was a discourse at the LAC about least-cost paths (LCP):**

- Opinion 1: LCP follow a **capitalistic** ideology since their aim is
  the best optimization. They are blind for post-processual ideas
- Opinion 2: LCP do optimize a certain path. But the rules, how they
  do it, is defined by **parameters** chosen for the model. Hence,
  they are able to grasp post-processual ideas

## Distance analyses in geographic space

Basic work about least-cost distances and path analyses (usefulness, applicability, problems,...) was/is conducted quite excessively by Irmela Herzog (https://bodendenkmalpflege-lvr.academia.edu/IrmelaHerzog)

## Distance analyses in geographic space {.columns-2}

What does it mean?

```{r}
## km/h
tobler1993a <- function(s){
    6 * exp(-3.5 * abs(s + 0.05))
}
## m/min
tobler1993b <- function(s){
    0.36 * exp(-3.5 * abs(s + 0.05))
}   
```

```{r echo=FALSE, fig.width = 5}
plot(tobler1993a,
     xlim = c(-1,1),
     main = "Slope-dependent cost function (Tobler)",
     xaxt="n",yaxt="n",
     xlab = "",ylab = "")

mtext(side = 1, text = "slope", line = 2)
mtext(side = 2, text = "speed (km/h)", line = 2)

axis(2, mgp=c(3, .5, 0))
axis(1, mgp=c(3, .5, 0))

abline(v = 0, lty = 2)
```

## Distance analyses in geographic space

Technically speaking, there are two different ways how least-cost paths are created

1. create a cost raster
2. accumulate the costs from one of your points of interest
3. "walk" from a target point along the accumulated raster to your starting point (like water flows into a sink)

In `GrassGIS`:

- cost raster
- `r.cost`/`r.walk`
- `r.drain`

## Distance analyses in geographic space

Technically speaking, there are two different ways how least-cost paths are created

In `R`:

1. create cost raster
2. create matrix of cell connections
3. find "cheapest" path between points along cell connections

`gdistance` in `R` a package by Jacob van Etten (https://cran.r-project.org/web/packages/gdistance)

## Distance analyses in geographic space

What are the necessary steps?

1. Create cost surface

```{r fig.width=5, echo=FALSE}
library(png)
library(grid)
img <- readPNG('assets/img/pic_daniel_transition_raster.png')
grid.raster(img)
```

(https://cran.r-project.org/web/packages/gdistance/vignettes/gdistance1.pdf)

## Distance analyses in geographic space

What are the necessary steps?

1. Create cost surface
2. Create transition matrix, i.e. matrix recording connections between cells

```{r fig.width=5, echo=FALSE}
library(png)
library(grid)
img <- readPNG('assets/img/pic_daniel_transition.png')
grid.raster(img)
```

(https://cran.r-project.org/web/packages/gdistance/vignettes/gdistance1.pdf)


## Distance analyses in geographic space

What are the necessary steps?

1. Create cost surface
2. Create transition matrix, i.e. matrix recording connections between cells
3. Transition matrix is filled with conductance, rather than resistance values; hence unconnected cells have value 0 (= no conductance); memory efficient

## Distance analyses in geographic space

What are the necessary steps?

1. Create cost surface
2. Create transition matrix, i.e. matrix recording connections between cells
3. Transition matrix is filled with conductance, rather than resistance values; hence unconnected cells have value 0 (= no conductance); memory efficient
4. Geocorrection of values

## Distance analyses in geographic space

What are the necessary steps?

1. Create cost surface
2. Create transition matrix, i.e. matrix recording connections between cells
3. Transition matrix is filled with conductance, rather than resistance values; hence unconnected cells have value 0 (= no conductance); memory efficient
4. Geocorrection of values
5. Identify adjacent cells

## Distance analyses in geographic space

What are the necessary steps?

1. Create cost surface
2. Create transition matrix, i.e. matrix recording connections between cells
3. Transition matrix is filled with conductance, rather than resistance values; hence unconnected cells have value 0 (= no conductance); memory efficient
4. Geocorrection of values
5. Identify adjacent cells
6. Calculate walking speed for adjacent cells

## Distance analyses in geographic space

What are the necessary steps?

1. Create cost surface
2. Create transition matrix, i.e. matrix recording connections between cells
3. Transition matrix is filled with conductance, rather than resistance values; hence unconnected cells have value 0 (= no conductance); memory efficient
4. Geocorrection of values
5. Identify adjacent cells
6. Calculate walking speed for adjacent cells
7. Geocorrect speed values

## Distance analyses in geographic space

What are the necessary steps?

1. Create cost surface
2. Create transition matrix, i.e. matrix recording connections between cells
3. Transition matrix is filled with conductance, rather than resistance values; hence unconnected cells have value 0 (= no conductance); memory efficient
4. Geocorrection of values
5. Identify adjacent cells
6. Calculate walking speed for adjacent cells
7. Geocorrect speed values
8. Find path using least-cost ("as the wolf runs") or random walk ("drunkard's walk") algorithm

## Distance analyses in geographic space {.columns-2}

First we calculate the slope, as this is used as cost

```{r message=FALSE}
library(raster)
dem <- raster("results/srtm.tif")
dem <- aggregate(x = dem, fact = 10)
slope <- terrain(x = dem,
                 opt = "slope",
                 neighbors = 8,
                 unit = "degrees")
slope <- crop(x = slope,
              y = extent(slope)-1000
              )
```

```{r echo=FALSE, fig.width = 5}
plot(slope)
```

## Distance analyses in geographic space

Now we create the transition object and geocorrect it

```{r message=FALSE}
library(gdistance)
slope.tran <- transition(x = slope,
                         transitionFunction = mean,
                         directions = 8,
                         symm = TRUE
                         )
slope.geo <- geoCorrection(x = slope.tran, scl = TRUE)
```

To see how the results look like we create a raster stack for easy plotting

```{r message=FALSE}
slopes <- stack(slope,
                raster(slope.tran),
                raster(slope.geo))
names(slopes) <- c("Slope","Transition","Geocorrection")
```

## Distance analyses in geographic space

To see how the results look like we create a raster stack for easy plotting

```{r message=FALSE, fig.width = 10}
plot(slopes, nr = 1)
```

## Distance analyses in geographic space

Now, the final step. Calculate Tobler's hiking speed, geocorrect again and calculate a shortest path

```{r message=FALSE}
adj <- adjacent(x = slope,
                cells=1:ncell(slope),
                pairs=TRUE,
                directions=8)
speed <- slope.geo
speed[adj] <-  6 * exp(-3.5 * abs(slope.geo[adj] + 0.05))

speed.geo <- geoCorrection(x = speed, scl=TRUE)

sp1 <- shortestPath(x = speed.geo,
                    origin = sites@coords[4,],
                    goal = sites@coords[50,],
                    output = "SpatialLines")
```

## Distance analyses in geographic space

Now, the final step. Calculate Tobler's hiking speed, geocorrect again and calculate a shortest path

```{r, fig.width=10}
plot(raster(speed.geo))
lines(sp1)
```

## Distance analyses in geographic space {.columns-2}

What about the "drunkard's walk"? 

```{r message=FALSE, fig.width = 5}
p1 <- passage(x = speed.geo,
              origin = sites@coords[4,],
              goal = sites@coords[50,]#,
              ##theta = .005
              )              
              
hs <-  hillShade(
    slope = terrain(x = dem*10, opt = "slope"),
    aspect = terrain(x = dem, opt = "aspect"),
    angle = 40, direction = 270
)

plot(hs, col=grey(0:100/100), legend=FALSE)
plot(p1, alpha = .5, add=TRUE)#; lines(sp1)
```

## Distance analyses in geographic space

What about the idea of a **prehistoric autobahn**? Run the code and get some coffee (most probably)

```{r message=FALSE}
xy <- data.frame(X = sites@coords[,1],
                 Y = sites@coords[,2])

rows.xy <- row.names(xy)
sa.rows.xy <- sample(row.names(xy), length(xy$X)/2)
starts <- subset(xy, rows.xy %in% sa.rows.xy)
goals <- subset(xy, !(rows.xy %in% sa.rows.xy))
sg.passages <- brick(slope)

for(i in 1:length(starts$X)) {
    s <- c(starts$X[i],starts$Y[i])
    z <- c(goals$X[i],goals$Y[i])
    sg.passages[[i]] <- passage(x = speed.geo, origin = s, goal = z)
    cat("iteration ", i, " of ", length(starts$X),"\n")
}

sg.passages.sum <- sum(sg.passages)
```

## Distance analyses in geographic space

```{r echo=FALSE, fig.width=10}
plot(hs, col=grey(0:100/100), legend=FALSE)
plot(sg.passages.sum,
     col = colorRampPalette(colors = c("white","red"))(50),
     alpha = .8,
     add=TRUE)
points(sites,
       pch = 19,
       col = rgb(red = 0, green = 0, blue = 0, alpha = .2))
```

